---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd (this file) -->

```{r setup, echo = FALSE, warning = FALSE}
chooseCRANmirror(graphics = FALSE, ind = 1)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Connectedness
An R package to 1) measure the connectedness across management units using genomic or pedigree data and 2) assess the connectedness between training and testing sets in genomic prediction. 

## Installation
Connectedness is currently available in Github, and can be installed with devtools package: 

1. Install `devtools` package from CRAN. 
```{r, warning = FALSE, message = FALSE, results = 'hide', eval = FALSE}
install.packages("devtools")
```

2. Load the `devtools` package.
```{r, warning = FALSE, results = 'hide', eval = FALSE}
library(devtools)
```

3. Install `Connectedness` package from Github.
```{r, warning = FALSE, results = 'hide', message = FALSE, eval = FALSE}
install_github('HaipengU/Connectedness', build_opts = c("--no-resave-data", "--no-manual"))
```

4. Load `Connectedness` package. 
```{r, warning = FALSE, results = 'hide', eval = FALSE}
library(Connectedness)
```

5. View `Connectedness` vignette.
```{r, warning = FALSE, results = 'hide', eval = FALSE}
vignette('Connectedness', package = 'Connectedness')
```

## Documentation

[Vignette](https://haipengu.github.io/Rmd/Vignette.html)
