---
title: "Connectedness Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Connectedness Package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
**Prediction Error Variance (PEV)**

Genetic connectedness is typically defined as a function of the inverse of the coefficient matrix (i.e., Prediction error variance (PEV)). The PEV can be obtained from Henderson's mixed model equations (MME). Assume a standard linear mixed model $\mathbf{y} = \mathbf{Xb} + \mathbf{Zu} + \boldsymbol{\epsilon}$, where $\mathbf{y}$, $\mathbf{b}$, $\mathbf{u}$ and $\boldsymbol{\epsilon}$ refer to a vector of phenotypes, effects of management units, random additive genetic effects and residuals, respectively. The $\mathbf{X}$ and $\mathbf{Z}$ are incidence matrices which associate management units effects and individuals to phenotypes, respectively. The inverse of the MME coefficient matrix derived from this model is 
\begin{align*}
\mathbf{C}^{-1} &= 
\begin{bmatrix}
 \mathbf{X'X} &      \mathbf{X'Z} \\
    \mathbf{Z'X} &     \mathbf{Z'Z} +  \mathbf{K}^{-1} \lambda  
 \end{bmatrix}^{-1} \\
&= 
\begin{bmatrix}
    \mathbf{C}^{11} & \mathbf{C}^{12} \\
     \mathbf{C}^{21} & \mathbf{C}^{22}
\end{bmatrix} 
\end{align*}
where $\lambda$ is the ratio of variance components $\frac{\sigma^2_{\epsilon}}{\sigma^2_u}$. With this, the PEV matrix can be computed as 
\begin{align*}
\text{PEV} &= (\mathbf{Z'MZ} + \mathbf{K}^{-1} \lambda)^{-1} \sigma^2_{\epsilon} \\
&= \mathbf{C}^{22} \sigma^2_{\epsilon}, 
\end{align*}
where $\mathbf{M} = \mathbf{I} - \mathbf{X}(\mathbf{X}' \mathbf{X})^{-}\mathbf{X}'$ is the absorption (projection) matrix for fixed effects. 

**Variance of Estimates of Management Units Effects(Var($\widehat{B}$))**

The derivation of PEV results in a high computational demanding for the high-dimensional dataset. Alternatively, Kennedy and Trus proposed a quantification of connectedness with variance of management units estimates, which is 
$$ \text{Var}({\widehat{B})} = [\mathbf{X}' \mathbf{X} - \mathbf{X}' \mathbf{Z}(\mathbf{Z'Z} + \mathbf{K}^{-1} \lambda)^{-1}\mathbf{Z}' \mathbf{X}]^{-1}\sigma^2_{\epsilon}$$


## Connectedness Statistics
### Prediction Error Variance of Difference (PEVD)
The PEVD measures the prediction error variance difference of breeding values between individuals from different management units. The PEVD between two individuals of $i$ and $j$ can be expressed as
\begin{align*} 
 \text{PEVD}(\hat{g_i} - \hat{g_j})  &= [ \text{PEV}(\hat{g_i}) + \text{PEV}(\hat{g_j}) - 2  \text{PEC}(\hat{g_i}, \hat{g_j})]  \\
 &= (\mathbf{C}^{22}_{ii} - \mathbf{C}^{22}_{ij} - \mathbf{C}^{22}_{ji} + \mathbf{C}^{22}_{jj}) \sigma^2_{\epsilon} \\
&= (\mathbf{C}^{22}_{ii}  + \mathbf{C}^{22}_{jj} - 2\mathbf{C}^{22}_{ij}) \sigma^2_{\epsilon} , 
\end{align*}
where PEC$_{ij}$ is the off-diagonal element of the PEV matrix which indicates the prediction error covariance or covariance between errors of genetic values. A summary of PEVD across any pair of management units under a contrast form is \begin{align*}
\text{PEVD}(\mathbf{x}) &= \mathbf{x}' \mathbf{C}^{22} \mathbf{x} \sigma^2_{\epsilon},
\end{align*} 
where the sum of elements in $\mathbf{x}$ equals to zero. 

**Example**

Use `?cattle` to learn more about this cattle dataset.
```{r warning = FALSE, message = FALSE, results = 'hide',eval = FALSE}
# Load cattle dataset from Connectedness package.
library(Connectedness)
data(GCcattle)

# Construct numerator relationship matrix (NRM)
library(kinship2)
A <- 2*kinship(cattle.pheno$Progeny, cattle.pheno$Sire, cattle.pheno$Dam)

# X matrix of least connected design
X_fixed <- model.matrix(~ -1 + factor(cattle.pheno$Management-Unit)) 

# Estimate variance components
varcomp <- VarComp_est(cattle.pheno$Phen, A, X_fixed)
varcomp$sigma2a # variance of additive genetic effects
varcomp$sigma2e # varaince of residual

# Calculate PEVD 
pevd_connectedness <- GC(Kmatrix = A, Xmatrix = X_fixed, sigma2a = varcomp$sigma2a, 
                        sigma2e = varcomp$sigma2e, MUScenario = as.factor(cattle.pheno$Management-Unit), 
                        statistic = 'PEVD', NumofMU = 'Pairwise')
pevd_connectedness # elements of Pairwise PEVD
```


### Coefficient of Determination (CD)
The coefficient of determination (CD) measures the precision of the estimates breeding values. The pair-wise CD between individual $i$ and $j$ is defined as
\begin{align*}
\text{CD}_{ij} &= \frac{var(\mathbf{g}) - var(\mathbf{g}|\mathbf{\hat{g}})}{var(\mathbf{g})} \\
&= 1 - \frac{var(\mathbf{g}|\mathbf{\hat{g}})}{var(\mathbf{g})} \\
&= 1 - \lambda \frac{\mathbf{C}^{22}_{ii} + \mathbf{C}^{22}_{jj} - 2\mathbf{C}^{22}_{ij} }{\mathbf{K}_{ii} + \mathbf{K}_{jj} - 2\mathbf{K}_{ij}}. 
\end{align*}
A contrast notation of summary CD between any pair of management units is 
\begin{align*}
\text{CD}(\mathbf{x}) &= 1 - \frac{var(\mathbf{x}'\mathbf{g}|\mathbf{\hat{g}})}{var(\mathbf{x}'\mathbf{g})} \\
&= 1- \lambda \frac{ \mathbf{x}' \mathbf{C}^{22} \mathbf{x} }{ \mathbf{x}' \mathbf{K} \mathbf{x}},
\end{align*}
where $\mathbf{x}$ is a contrast vector defined above and $\mathbf{K}$ is relationship matrix. 

**Example**
```{r eval = FALSE}
# CD with least connected design
cd_connectedness <- GC(Kmatrix = A, Xmatrix = X_fixed, sigma2a = varcomp$sigma2a, 
                        sigma2e = varcomp$sigma2e, MUScenario = as.factor(cattle.pheno$Management-Unit), 
                        statistic = 'CD', NumofMU = 'Pairwise')

cd_connectedness # elements of pairwise CD
```

### Prediction Error Correlation (r)
The pairwise r statistic between individual $i$ and $j$ can be derived from PEV matrix to a predictive error correlation matrix as 
\begin{align*}
\text{r}_{ij} = \frac{\text{PEC}(\hat{g_i}, \hat{g_j})}{\sqrt{\text{PEV}(\hat{g_i})  \text{PEV}(\hat{g_j})}}.
\end{align*}

- **Flock connectedness**
One of the summary statistics of r is flock connectedness, which takes the average of PEV matrix by management units followed by a ratio to r. The Flock r between two management units $i'$ and $j'$ is given by  
\begin{align*}
\text{r}_{i'j'} &= \frac{ 1/n_{i'} \sum \text{PEC}_{i' j'} 1/n_{j'}}{   \sqrt{ (1/n_{i'})^2 \sum \text{PEV}_{i' i'} \cdot (1/n_{j'})^2 \sum \text{PEV}_{j' j'} }   } \\
&= \frac{ \sum \text{PEC}_{i' j'} }{   \sqrt{  \sum \text{PEV}_{i' i'} \cdot  \sum \text{PEV}_{j' j'} }   }, 
\end{align*}
where $n_{i'}$ and $n_{j'}$ indcate the number of individuals in units $i'$ and $j'$, respectively. 
- **Individual connectedness**
The summary of individual connectedness calculates pairwise r for all pair of individuals followed by averaging all r across management units: 
\begin{align*}
\text{r}_{i'j'} = \overline{\sum{\frac{\text{PEC}(\hat{g_{i'}}, \hat{g_{j'}})}{\sqrt{\text{PEV}(\hat{g_{i'}})  \text{PEV}(\hat{g_{j'}})}}}}.
\end{align*}

**Example**
```{r eval = F}
# Flock connectedness
flock_r <- GC(Kmatrix = A, Xmatrix = X_fixed, sigma2a = varcomp$sigma2a, 
                  sigma2e = varcomp$sigma2e, MUScenario = as.factor(cattle.pheno$Management-Unit), 
                  statistic = 'r2', NumofMU = 'Pairwise')
flock_r # contents of flock r

# Individual connectedness
individual_r <- GC(Kmatrix = A, Xmatrix = X_fixed, sigma2a = varcomp$sigma2a, 
                  sigma2e = varcomp$sigma2e, MUScenario = as.factor(cattle.pheno$Management-Unit), 
                  statistic = 'r1', NumofMU = 'Pairwise')
individual_r # contents of flock r
```

### Variance of Differences in Management Units Effects (VED)
The VED statistic estimates connectedness with the variance of differences of management units effects. The VED between two units $i'$ and $j'$ is defined as
\begin{align*}
\text{VED}_{i'j'} &=
\text{Var}(\widehat{B})_{i'i'} + \text{Var}(\widehat{B})_{j'j'} - 2\text{Cov}(\widehat{B})_{i'j'},
\end{align*}
where Var$(\widehat{B})_{i'i'}$, Var$(\widehat{B})_{j'j'}$ and Cov$(\widehat{B})_{i'j'}$ indicate variance of management unit $i'$, $j'$ and covariance between two units, respectively. 

**Example**
```{r eval=FALSE}
# VED of least connected design
ved_connectedness <- GC(Kmatrix = A, Xmatrix = X_fixed, sigma2a = varcomp$sigma2a, 
                  sigma2e = varcomp$sigma2e, MUScenario = as.factor(cattle.pheno$Management-Unit), 
                  statistic = 'VED', NumofMU = 'Pairwise')
ved_connectedness # elements of VED
```


### Connectedness Rating (CR)
The CR approximates flock connectedness by replacing PEV matrix with the variance of differences of management units estimates. The CR of two management units $i'$ and $j'$ is given by
\begin{align*}
\text{CR}_{i'j'} &=
\frac{\text{Cov}(\widehat{B})_{i'j'} }{ \sqrt{ \text{Var}(\widehat{B})_{i'i'} \cdot  \text{Var}(\widehat{B})_{j'j'}}}.
\end{align*}

**Example**
```{r eval=FALSE}
cr_connectedness <- GC(Kmatrix = A, Xmatrix = X_fixed, sigma2a = varcomp$sigma2a, 
                  sigma2e = varcomp$sigma2e, MUScenario = as.factor(cattle.pheno$Management-Unit), 
                  statistic = 'CR', NumofMU = 'Pairwise')
cr_connectedness # elements of CR
```
## Continue...
## Authors
- Haipeng Yu (<haipengyu@vt.edu>)
- Mehdi Momen(<momenmehdi@vt.edu>)
- Gota Morota (<morota@vt.edu>)